% Auf ungerader Seite starten
\cleardoublepage

\chapter{Introduction}
\label{Chapter:Introduction}
Aircraft engines are among the most safety-critical and economically significant components in modern aviation. A single unplanned engine failure during flight can endanger lives. On the ground, unscheduled maintenance events disrupt airline operations, strand passengers and generate costs that cascade through the network, from replacement parts and hangar time to delayed flights and rebooking. Engine maintenance, repair and overhaul (MRO) accounts for a large share of an airline's total operating expenditure and cost difference between a well-timed maintenance intervention and an emergency shutdown can be orders of magnitude. 

For most of aviation's history, maintenance has followed either a reactive or a scheduled approach. Reactive maintenance is unacceptable for flight-critical components. Scheduled maintenance (inspect or replace at fixed intervals regardless of actual condition) is safer but expensive, because components are often serviced long before they need to be. Parts with remaining useful life are discarded, inspection windows are conservatively short and aircraft spend more time in hangars than necessary. 

Condition-based and predictive maintenance aim to do better \citep{Jardine2006}. Rather than relying on fixed schedules, these approaches use operational data (sensor readings collected during normal operation, also called Snapshots) to estimate how much useful life a component has left. If the remaining useful life (RUL) of an engine can be predicted with reasonable accuracy, maintenance can be scheduled for when it is actually needed: late enough to extract most of a component's service life, early enough to avoid failure. The potential savings are significant, both in direct maintenance costs and in improved aircraft availability. For safety-critical systems like turbofan engines, the added benefit is that degradation can be tracked continuously rather than assessed at discrete inspection intervals, reducing the chance that a developing fault goes undetected between scheduled checks.

Estimating RUL from sensor data is, at its core, a prediction problem. An engine's condition at any point in time is reflected in the readings from its sensors: temperatures, pressures, speeds, vibration levels. The task is to take a history of these readings and predict how many more operational cycles the engine can sustain before failure. Over the past decade, machine learning models have become the dominant approach to this problem \citep{Li2018,Zheng2017}. They learn the mapping from sensor histories to RUL directly from run-to-failure data, without requiring explicit physical models of the degradation process. On standard benchmark datasets such as C-MAPSS \citep{Saxena2008}, these data-driven methods now achieve prediction errors that would have been difficult to reach with earlier approaches.

But prediction accuracy alone is not sufficient for adoption in safety-critical applications. A model that predicts an engine has 50 cycles of remaining life is only useful if the engineers and decision-makers who act on that prediction can understand why the model arrived at that number. Which sensor readings drove the prediction? Is the model responding to a genuine degradation signal, or to a statistical artefact in the training data? Would the prediction change if a particular sensor were faulty or miscalibrated? These questions matter because the consequences of a wrong prediction in aviation maintenance are not symmetric, overestimating RUL risks a failure in service, underestimating it wastes resources and availability. Without insight into the model's reasoning, there is no way to assess whether a specific prediction is trustworthy in a specific operational context.

This is the problem that eXplainable Artificial Intelligence (XAI) addresses. XAI encompasses methods that make the internal workings or outputs of machine learning models more interpretable to humans \citep{BarredoArrieta2020}. Some methods explain individual predictions, why the model gave \textit{this} answer for \textit{this} input. Others characterise global behaviour, which input features matter most across all predictions and how they influence the output. The methods differ in their assumptions, their computational mechanisms and the kinds of questions they answer. A gradient-based method traces importance through the model's own parameters. A perturbation based method observes how predictions change when inputs are modified. A surrogate-based method fits a simpler, interpretable model to approximate the complex one locally. Each perspective captures something real about the model's behaviour, but none captures everything.

This thesis takes the multi-method approach. Eight feature importance methods are applied to a 1D Convolutional Neural Network (CNN) trained for RUL prediction on the NASA C-MAPSS turbofan engine dataset \citep{Saxena2008}. The methods span the major categories of XAI (gradient-based, perturbation-based, Shapley-value-based, marginal-effect based and counterfactual) and are compared systematically across multiple training configurations. The analysis identifies which sensors the model considers most important, evaluates how consistently different methods agree on that ranking, examines how feature importance is distributed across the temporal dimension of  the input and tests whether the model can maintain its predictive performance using only the most important sensors.

The thesis is structured as follows. Chapter~\ref{Chapter:Related_work} reviews related work on data-driven RUL prediction and explainability in prognostics, identifying the gap that this thesis addresses. Chapter~\ref{Chapter:Background} introduces the theoretical foundations of the eight feature importance methods used. Chapter~\ref{Chapter:Methodology} describes the dataset, model architecture, preprocessing pipeline and experimental setup. Chapter~\ref{Chapter:Results} presents the results, from the verification of the implementations through the comparative analysis, temporal attribution, reduced sensor experiments and the discussion of their implications. Chapter~\ref{Chapter:Conclusion} summarises the findings and outlines directions for future work.


